\documentclass{article}

\usepackage{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}

\title{MoleculeLens: An Explainable Thin Bridge for Drug--Text Alignment and Retrieval}

\author{%
Anonymous Authors \\
Anonymous Institution \\
\texttt{anonymous@anonymous.edu}
}

\begin{document}

\maketitle

\begin{abstract}
Aligning drug representations with mechanism-of-action text is useful for retrieval, hypothesis generation, and biomedical search, but current benchmarks can overstate progress when lexical cues or train--test overlap make retrieval artificially easy. This paper presents \textit{MoleculeLens}, a scaffold-split extension of Thin Bridges for drug--text alignment on ChEMBL mechanism pairs. MoleculeLens uses frozen text representations, ECFP4 molecular fingerprints, and lightweight learned projections trained with a contrastive objective. The evaluation is built from 2,699 approved-drug mechanism pairs from ChEMBL and a reproducible Bemis--Murcko scaffold split with 2,019 train, 245 validation, and 435 held-out test examples. On the held-out scaffold test set, MoleculeLens reaches Recall@1 of 0.179 and MRR of 0.270, outperforming MolPrompt (0.129 Recall@1, 0.256 MRR), KV-PLM (0.055 Recall@1, 0.121 MRR), and a zero-shot Graphormer baseline (0.002 Recall@1, 0.015 MRR). Removing drug names from the text reduces MoleculeLens Recall@1 from 0.179 to 0.106, a smaller relative drop than the earlier full-gallery memorization-heavy Thin Bridges setting. These results suggest that thin contrastive bridges remain competitive under scaffold-disjoint evaluation, while also revealing that leakage-sensitive text fields still materially affect retrieval performance.
\end{abstract}

\section{Introduction}

Drug--text alignment asks whether a molecular representation and a natural-language description of mechanism or therapeutic function can be mapped into a common retrieval space. A robust solution would support mechanism-based search, structured pharmacology retrieval, and retrieval-augmented drug discovery workflows. In practice, however, strong retrieval scores can emerge for the wrong reasons. Drug names can leak identity into text, random or global train--test splits allow memorization, and benchmark choices can conflate semantic alignment with surface-form matching.

Our earlier Thin Bridges formulation showed that a lightweight contrastive bridge can strongly align drug structure and text. The main weakness of that earlier setting is that its strongest result used a full-gallery setup with train=test, which is useful as a memorization reference but not as the central scientific claim. The main-conference version therefore shifts emphasis to a stricter scaffold-split evaluation and repositions the scaffold model as \textit{MoleculeLens}. The goal is not to claim that the problem is solved, but to show that a lightweight bridge remains competitive when molecules in the test set have unseen Bemis--Murcko scaffolds.

This draft makes three concrete contributions. First, it defines a reproducible ChEMBL retrieval benchmark based on approved drug--mechanism pairs with a scaffold-disjoint split. Second, it evaluates the thin-bridge approach against a broader set of retrieval baselines, including MolPrompt, KV-PLM, and a Graphormer-based zero-shot baseline. Third, it revisits the drug-name leakage question under the scaffold split and shows that the scaffold-disjoint model is substantially less brittle than the earlier global memorization-heavy setting, although it is still not immune to lexical shortcutting.

\section{Related Work}

\paragraph{Molecule--text representation learning.}
Recent work has explored joint learning over molecular structures and natural language, including retrieval-oriented models such as Text2Mol~\citep{edwards2021text2mol}, multimodal pretrained models such as KV-PLM~\citep{zeng2022kvplm}, and broader molecular benchmark suites such as MoleculeNet~\citep{wu2018moleculenet}. These systems differ in molecular modality, text domain, and supervision strategy, but they share a common assumption that chemically meaningful structure--text correspondences can be captured in a shared space.

\paragraph{Graph and language baselines.}
Transformer architectures for graphs have become standard baselines for molecular prediction, with Graphormer as a representative example~\citep{ying2021graphormer}. In parallel, text encoders pretrained on scientific or biomedical corpora such as SciBERT~\citep{beltagy2019scibert} have influenced biomedical retrieval systems and text-conditioned molecular models. Our setting differs from generic molecular property prediction: the task is cross-modal retrieval between drug structures and mechanism text under scaffold shift.

\paragraph{Evaluation leakage and scaffold splits.}
Scaffold-based evaluation remains a standard strategy for reducing overoptimistic estimates in molecular machine learning~\citep{bemis1996frameworks}. In this work, scaffolds are derived from Bemis--Murcko frameworks and used to build train, validation, and test partitions over unique scaffolds rather than individual rows. The molecular side uses ECFP4 fingerprints~\citep{rogers2010ecfp} as a simple, strong structural representation. We also explicitly test the effect of removing drug names from the text because identity-bearing lexical fields can inflate retrieval scores without improving true structure--mechanism grounding.

\section{Benchmark and Task}

\subsection{ChEMBL Drug--Mechanism Pairs}

The benchmark is constructed from ChEMBL~\citep{zdrazil2024chembl}. Starting from the mechanism endpoint, we keep approved drugs only by filtering for \texttt{max\_phase=4} and molecular mechanisms with valid mechanism text. Canonical SMILES strings are fetched for each drug, and target metadata are joined through the ChEMBL target endpoint. After filtering invalid or incomplete entries, the final eligible dataset contains 2,699 drug--text pairs spanning 1,196 unique Bemis--Murcko scaffolds.

Each example contains a drug structure and a text description. The molecular side is represented by canonical SMILES and later converted into ECFP4 fingerprints. The text side is a concatenation of mechanism of action, target name, action type, and drug name:
\begin{equation}
\texttt{text\_rich} = \texttt{mechanism} + \texttt{target} + \texttt{action} + \texttt{drug}.
\end{equation}
This formulation is intentionally rich because the benchmark is designed both to measure retrieval accuracy and to expose sensitivity to lexical cues such as explicit drug-name mentions.

\subsection{Retrieval Setup}

The task is bidirectional retrieval: given a text query, retrieve the matching drug structure from a gallery, and given a structure query, retrieve the matching text. We report full-gallery Recall@1, MRR, Recall@5, and Recall@10 as the primary metrics. We also retain T-choose-one retrieval as a secondary diagnostic because it has appeared in earlier experiments, but it is not the main ranking criterion in this draft.

\subsection{Scaffold Split}

The main evaluation uses a saved scaffold manifest built over unique Bemis--Murcko scaffolds with seed 0. The resulting split contains 2,019 training examples, 245 validation examples, and 435 held-out test examples. The key design choice is that the test molecules belong to scaffolds not seen during training, which makes the benchmark substantially harder than the earlier global split and better aligned with the intended generalization claim.

\section{Method}

\subsection{Thin Contrastive Bridge}

MoleculeLens keeps the central simplicity of Thin Bridges. Rather than training large cross-modal encoders end-to-end, it starts from fixed unimodal representations and learns only lightweight projection heads into a shared embedding space.

For molecules, we compute ECFP4 fingerprints with radius 2 and 2,048 bits~\citep{rogers2010ecfp}. For text, we encode \texttt{text\_rich} with a frozen biomedical sentence encoder and use the CLS-style pooled representation. Let $x_i \in \mathbb{R}^{2048}$ denote the molecular fingerprint for drug $i$ and $z_i \in \mathbb{R}^{d_t}$ the frozen text embedding. MoleculeLens learns linear projections
\begin{equation}
b_i^{(m)} = W_m x_i, \qquad b_i^{(t)} = W_t z_i,
\end{equation}
where both projected embeddings live in a 256-dimensional shared space and are $L_2$ normalized before scoring.

\subsection{Training Objective}

Training uses a symmetric contrastive loss over matched drug--text pairs. Given a minibatch of projected text embeddings $B^{(t)}$ and molecular embeddings $B^{(m)}$, similarities are computed with a temperature-scaled cosine form:
\begin{equation}
s_{ij} = \frac{\langle \hat{b}_i^{(t)}, \hat{b}_j^{(m)} \rangle}{\tau}.
\end{equation}
The loss combines bidirectional InfoNCE with two modifications used in the current codebase: same-target weighting for hard negatives that share a target label, and a margin term against the hardest same-target negative in the minibatch. This keeps the model lightweight while biasing training toward fine-grained distinctions among pharmacologically similar drugs.

\section{Experimental Setup}

\subsection{Baselines}

We compare MoleculeLens against four families of baselines already present in the experimental repository.

\begin{itemize}
\item \textbf{MoleculeSTM (zero-shot):} an earlier molecular structure--text model evaluated from prior head-to-head runs.
\item \textbf{Graphormer (zero-shot):} a graph-based baseline evaluated on the aligned scaffold split.
\item \textbf{MolPrompt:} a trained retrieval baseline on the same held-out ChEMBL-style split.
\item \textbf{KV-PLM:} a biomedical language-model baseline that encodes both structure strings and text in a shared representation~\citep{zeng2022kvplm}.
\end{itemize}

We also retain the earlier \textbf{Thin Bridges (Global)} result as a reference point. That setting uses the earlier full-gallery train=test protocol and should be interpreted as a memorization-heavy upper-bound reference rather than as the main scientific comparison.

\subsection{Implementation Notes}

The current scaffold model uses batch size 512, learning rate $10^{-3}$, weight decay $10^{-4}$, temperature 0.07, margin 0.1, and 100 training epochs. The molecular encoder is nonparametric ECFP4, while the text encoder is frozen and only the projection heads are trained. A separate no-drug-name condition removes the \texttt{Drug: X.} span from the text before encoding.

\section{Results}

\subsection{Primary Retrieval Comparison}

Table~\ref{tab:main-results} shows the main comparison. MoleculeLens achieves the strongest Recall@1 and MRR among the held-out scaffold models, improving on MolPrompt by roughly five Recall@1 points and on KV-PLM by more than twelve Recall@1 points. The zero-shot Graphormer baseline performs near chance on this retrieval task. These results suggest that the thin-bridge design remains competitive when evaluated on unseen scaffolds rather than on the earlier full-gallery setup.

\begin{table}[t]
\centering
\small
\caption{Primary retrieval comparison. Full-gallery rows are included for context; the held-out scaffold rows are the main scientific comparison.}
\label{tab:main-results}
\begin{tabular}{l l c c c c}
\toprule
Model & Eval Set & Recall@1 & MRR & Recall@5 & Recall@10 \\
\midrule
MoleculeSTM (zero-shot) & Full gallery ($N{=}2699$) & 0.089 & 0.171 & 0.243 & 0.337 \\
Graphormer (zero-shot) & Scaffold test ($N{=}435$) & 0.002 & 0.015 & 0.011 & 0.025 \\
Thin Bridges (Global) & Full gallery ($N{=}2699$) & 0.747 & 0.849 & 0.985 & 0.997 \\
\textbf{MoleculeLens (ours)} & Scaffold test ($N{=}435$) & \textbf{0.179} & \textbf{0.270} & 0.379 & 0.455 \\
MolPrompt & Scaffold test ($N{=}435$) & 0.129 & 0.256 & \textbf{0.400} & \textbf{0.531} \\
KV-PLM & Scaffold test ($N{=}435$) & 0.055 & 0.121 & 0.172 & 0.251 \\
\bottomrule
\end{tabular}
\end{table}

Two points matter for interpretation. First, the global Thin Bridges result remains much higher than every scaffold-disjoint result, which reinforces that the earlier setting should be viewed as a reference rather than a fair main comparison. Second, MolPrompt attains stronger Recall@10 than MoleculeLens, suggesting that MoleculeLens is better at placing the correct match near the top of the list but may still lose ground on broader candidate coverage.

\subsection{Drug-Name Leakage}

Table~\ref{tab:leakage-results} revisits the leakage question by removing explicit drug names from the text. As expected, all text-aware models degrade to some extent. The important comparison is between the old global Thin Bridges setup and the new scaffold model. The global setting drops by 80.5\% relative Recall@1, while MoleculeLens drops by 41.0\%. This does not eliminate leakage, but it suggests that scaffold-disjoint evaluation plus a lighter bridge produces a less brittle retrieval model than the earlier memorization-heavy setup.

\begin{table}[t]
\centering
\small
\caption{Drug-name leakage ablation measured with Recall@1.}
\label{tab:leakage-results}
\begin{tabular}{l l c c c c}
\toprule
Model & Eval Set & With Drug & No Drug & Abs.\ Drop & \% Drop \\
\midrule
MoleculeSTM (zero-shot) & Full gallery ($N{=}2699$) & 0.089 & 0.014 & 0.074 & 83.8 \\
Graphormer (zero-shot) & Scaffold test ($N{=}435$) & 0.002 & 0.005 & -0.002 & -100.0 \\
Thin Bridges (Global) & Full gallery ($N{=}2699$) & 0.747 & 0.145 & 0.601 & 80.5 \\
\textbf{MoleculeLens (ours)} & Scaffold test ($N{=}435$) & \textbf{0.179} & \textbf{0.106} & \textbf{0.074} & \textbf{41.0} \\
MolPrompt & Scaffold test ($N{=}435$) & 0.129 & 0.071 & 0.057 & 44.6 \\
KV-PLM & Scaffold test ($N{=}435$) & 0.055 & 0.034 & 0.021 & 37.5 \\
\bottomrule
\end{tabular}
\end{table}

The Graphormer row is unstable because the absolute scores are already near chance, so its apparent negative drop should not be overinterpreted. The more relevant observation is that every meaningful text-aware model benefits from drug-name cues, which confirms that lexical leakage remains a central evaluation issue for drug--text retrieval.

\section{Discussion}

The current results support a narrower claim than the workshop version. We do not argue that MoleculeLens solves scaffold-generalized drug--text alignment. Instead, we argue that a lightweight bridge remains competitive under a substantially stricter evaluation protocol and offers a better balance between simplicity and performance than several stronger-looking alternatives.

At the same time, the benchmark still has clear limitations. The text field includes target and action metadata alongside mechanism descriptions, the dataset is modest in size, and the no-drug-name ablation shows that lexical identity cues remain influential. In addition, this draft does not yet include the interpretability analysis and broader ablation suite planned for the full submission. Those analyses are likely important for separating genuine mechanistic grounding from learned shortcut behavior.

\section{Conclusion}

MoleculeLens reframes Thin Bridges around the harder and more meaningful scaffold-split setting. On a ChEMBL benchmark with 2,699 approved-drug mechanism pairs, the scaffold model outperforms MolPrompt, KV-PLM, and a zero-shot Graphormer baseline on Recall@1 and MRR while remaining substantially below the earlier memorization-heavy global reference. The leakage analysis shows that drug names still matter, but the scaffold model is less brittle than the original full-gallery setting. This makes MoleculeLens a better foundation for the next stage of the project: interpretability, stronger ablations, and cleaner evidence for true structure--mechanism alignment.

\bibliographystyle{plainnat}
\bibliography{refs}

\newpage
\appendix

\section{Planned Additions for the Full Submission}

This draft establishes the core narrative and the main tables. The next revision should add:
\begin{itemize}
\item interpretability analyses for the scaffold model that motivate the ``MoleculeLens'' naming more directly;
\item non-neural baselines such as BM25 and ECFP4 Tanimoto search;
\item a cleaner appendix table for T-choose-one metrics and exact command lines;
\item a short reproducibility appendix describing artifact locations and split manifests.
\end{itemize}

\input{checklist}

\end{document}
